Staged Abstract Interpreters

### 2019-03-27 paper/10_intro.tex
mostly section 5

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections, which first identifies the binding-time of variables in
the program, i.e., they can be known whether statically or dynamically, then
evaluates the static part, and finally generates a residual program that solely
relies on the dynamic part. However, given an arbitrary program, precisely
analyzing its binding-times is hard in general. As an alternative and pragmatic
approach to specialization and partial evaluation, multi-stage programming (MSP
for short) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the stage
annotations (i.e., binding-time annotations) to be explicit from the
programmers. The stage annotations can be either syntactic (as in
MetaML/MetaOCaml, through quasi-quote, escape, etc.) or type-based \todo{cite}.
These staging annotations identify which part of the inputs should be
specialized. As a classical example, we use the power function and type-based
annotations to introduce the idea of MSP:

@@

where $e$ is the program of type @Expr@ and @arg@ is the input to program $e$,
and @eval@ produces a value.
Semantically, the interpreter satisfies $ \texttt{eval}(e)(arg) = [\![ e ]\!] arg$ for all
programs and inputs. Given a program $e_0$, by applying the first Futamura
projection, we may obtain a specialized interpreter
$\texttt{eval}_{\texttt{e0}}$ with respect to $e_0$ -- the so-called \textit{equivalent executable}. 
By definition of the interpreter, 
$\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $. If we generalize the
argument @arg@ to an environment (as well as an accompanying heap object, if for
the interest of using assignments) that maps free variables in $e$ to some
values, the interpreter is a standard environment-passing interpreter,
where the specialization through MSP is also applicable.

Roughly at the same time when Futamura projections was visioned in 1970s,
\citet{DBLP:conf/popl/CousotC77} proposed abstract interpretation as a
lattice-based semantic approach to construct sound static analysis, by
approximation of fix-points. However, constructing sound abstract interpreters
was considered abstruse and complicated for a long time.
Recent progress such as Abstracting Abstract Machines (AAM)
\cite{DBLP:conf/icfp/HornM10} uncovers a systematic method to derive sound
abstract interpreters from their concrete counterparts, where the soundness can
be easily established by examining the transformation of semantics. The AAM
approach also has been been applied to different variants of definitional
interpreters and abstract machines \cite{DBLP:journals/jfp/HornM12,
DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}.

@@

Now we have seen two orthogonal transformations of interpreters: by
specialization, we can refactor interpreters to code generators; by
approximation, we can refactor interpreters to static analyzers. As an
intellectual quest, we should naturally ask -- starting from an abstract
interpreter, \textit{can we derive a sound abstract interpreter while it
produces fast code that does the analysis?}
Conceptually, it is similar to the concrete setting: for an
abstract interpreter $\widehat{\texttt{eval}}: \texttt{Expr} \to
\widehat{\texttt{Env}} \to \widehat{\texttt{Value}}$, which takes a program, an abstract
environment and returns abstract values, we would like to specialize it with
respect to some program $e_0$ and produce a compiled analysis
$\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, such that $
\widehat{\texttt{eval}}_{\texttt{e0}}(\widehat{\rho}) =
\widehat{\texttt{eval}}(e_0)(\widehat{\rho})$, but the compiled one leaves the
abstract environment $\widehat{\rho}$ as remaining input.

@@

To enable such flexibility between different semantics, we combine generic 
programming and generative programming, and extensively use abstract type members 
over multiple abstractions, which can be later instantiated.

>>>>

Futamura projections \cite{Futamura1999, futamura1971partial} reveal the close
connection between compilers and interpreters. The first Futamura projection
specifically shows that specializing an interpreter with respect to the input
program yields an equivalent executable. Partial evaluation
\cite{DBLP:books/daglib/0072559} was the first proposed approach to realize
Futamura projections: it identifies the binding-time of variables in
the program (i.e., they can be known statically or dynamically),
evaluates the static part, and finally generates a residual program that solely
relies on the dynamic part. However, given an arbitrary program, precisely
analyzing its binding-times is hard in general. As an alternative and pragmatic
approach to specialization and partial evaluation, multi-stage programming (MSP
for short) \cite{taha1999multi, DBLP:conf/pepm/TahaS97} requires the staging
annotations (i.e., binding-time annotations) to be explicit from the
programmers. The staging annotations can be either syntactic (as in
MetaML/MetaOCaml, through quasi-quote, escape, etc.) or type-based \todo{cite}.
These staging annotations identify which part of the inputs should be
specialized. As a classical example, we use the power function and type-based
annotations to introduce the idea of MSP:

@@

where $e$ is the program of type @Expr@ and @arg@ is the input to program $e$,
and @eval@ produces a value.
Semantically, the interpreter satisfies $ \texttt{eval}(e)(arg) = [\![ e ]\!] arg$ for all
programs and inputs. Given a program $e_0$, by applying the first Futamura
projection, we may obtain a specialized interpreter
$\texttt{eval}_{\texttt{e0}}$ with respect to $e_0$ -- the so-called \textit{equivalent executable}. 
By definition of the interpreter, 
$\texttt{eval}_{\texttt{e0}}(arg) = [\![ e_0 ]\!] arg $. If we generalize the
argument @arg@ to an environment (as well as an accompanying heap object to
support mutable variables) that maps free variables in $e$ to some
values, the interpreter is a standard environment-passing interpreter,
where the specialization through MSP is also applicable.

Roughly at the same time when Futamura projections was envisioned in 1970s,
\citet{DBLP:conf/popl/CousotC77} proposed abstract interpretation as a
lattice-based semantic approach to construct sound static analysis, by
approximation of fix-points. However, constructing sound abstract interpreters
was considered abstruse and complicated for a long time.
Recent progress such as Abstracting Abstract Machines (AAM)
\cite{DBLP:conf/icfp/HornM10} uncovers a systematic method to derive sound
abstract interpreters from their concrete counterparts, where the soundness can
be easily established by examining the transformation of semantics. The AAM
approach also has been been applied to different variants of definitional
interpreters and abstract machines \cite{DBLP:journals/jfp/HornM12,
DBLP:conf/icfp/HornM10, DBLP:journals/pacmpl/DaraisLNH17}.

@@

Now we have seen two orthogonal transformations of interpreters: by
specialization, we can refactor interpreters to code generators; by
approximation, we can refactor interpreters to static analyzers. As an
intellectual quest, we should naturally ask -- starting from an abstract
interpreter, \textit{can we derive a sound abstract interpreter while it
produces optimized code that does the analysis?}
Conceptually, it is similar to the concrete setting: for an
abstract interpreter $\widehat{\texttt{eval}}: \texttt{Expr} \to
\widehat{\texttt{Env}} \to \widehat{\texttt{Value}}$, which takes a program, an abstract
environment and returns abstract values, we would like to specialize it with
respect to some program $e_0$ and produce a compiled analysis
$\widehat{\texttt{eval}}_{\texttt{e0}} : \widehat{\texttt{Env}} \to
\widehat{\texttt{Value}}$, such that $
\widehat{\texttt{eval}}_{\texttt{e0}}(\widehat{\rho}) =
\widehat{\texttt{eval}}(e_0)(\widehat{\rho})$, but the compiled one leaves the
abstract environment $\widehat{\rho}$ as remaining input.

@@

To enable such flexibility between different semantics, we combine generic 
programming and generative programming, and extensively use abstract type members 
over the following abstractions, which can be later instantiated.

### 2019-03-27 paper/30_stageconc.tex
mostly section 5

\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store]) => Rep[(Value, Store)]): Rep[Value]
  def close(ev: EvalFun)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[(Value, Store)] = {
      case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x); val ρ_* = ρ + (unit(x) → α); val σ_* = σ + (α + v)
        ev(e)(ρ_*)(σ_*)
    }; emit_compiled_clo(f)
  }
\end{lstlisting}

>>>>

\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store]) => Rep[(Value, Store)]): Rep[Value]
  def close(ev: EvalFun)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[(Value, Store)] = {
      case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x); val ρ_* = ρ + (unit(x) → α); val σ_* = σ + (α → v)
        ev(e)(ρ_*)(σ_*)
    }; emit_compiled_clo(f)
  }
\end{lstlisting}

### 2019-03-27 paper/50_stageabs.tex
section 4
mostly section 5

In the previous sections, we have seen an unstaged abstract interpreter and a
staged concrete interpreter, now we begin describing the implementation of their
confluence -- a staged abstract interpreter. We present a principled approach to
derive staged abstract interpreter from its unstaged version. One guiding
principle of our approach is that the code of the abstract semantics and the
code that optimizes should be separated. This it is an advantage of using
staging for abstract interpreters: the designer of the analysis has no need to
rewrite the analysis, and the performance improvement comes almost for free,
without any sacrifice of soundness or precision. Unsurprisingly, the staged
abstract interpreter we present in this section has the same abstract semantics
as the unstaged version we presented in Section~\ref{unstaged_abs}.

@@

In Section~\ref{stagedpoly_lat}, we exploited the higher-kinded type @R@ to
leave space for staging lattices, now we instantiate the type @R@ to @Rep@ and
still use powersets as an example to present its staged version.

\begin{lstlisting}
trait RepLattice[A] extends Lattice[A, Rep]
implicit def RepSetLattice[T:Typ]: RepLattice[Set[T]] = 
  new RepLattice[Set[T]] {
    lazy val bot: Rep[Set[T]] = Set[T]()
    lazy val top: Rep[Set[T]] = throw new NotImplementedError()
    def $\sqsubseteq$(l1: Rep[Set[T]], l2: Rep[Set[T]]): 
      Rep[Boolean] = l1 subsetOf l2
    def $\sqcup$(l1: Rep[Set[T]], l2: Rep[Set[T]]): 
      Rep[Set[T]] = l1 union l2
    def $\sqcap$(l1: Rep[Set[T]], l2: Rep[Set[T]]): 
      Rep[Set[T]] = l1 intersect l2
  }
\end{lstlisting}

The type parameter @T:Typ@ of @RepLattice@ requires that the elements of sets
can also be staged. Otherwise, without knowing how to stage the elements in the
set, we can not stage the set either. The methods defined operate on type
@Rep[Set[T]]@, thus the underlying implementation such as @union@ and
@intersect@ will be mapped to a node in the IR graph during staging and
eventually emitted in the generated code. Again, other structures such as maps
and tuples are implemented in a similar way.

@@

We have seen how to obtain a staged concrete semantics based on types, now we
take the same approach to obtain a staged abstract semantics. The basic
operations are largely kept the same as in the unstaged version, except the
types are changed to @Rep@. Besides, when we update the environment, the
identifier @x@ is known statically, but the environment map has type
@Rep[Map[Ident,Addr]]@, so we apply @lift@ to @x@ to turn it as a next-stage
value.

\begin{lstlisting}
trait RepAbsInterpOps extends Abstract with LMSOps {
  type R[+T] = Rep[T]
  val $\rho$0: Rep[Env] = Map[Ident, Addr]()
  val $\sigma$0: Rep[Store] = Map[Addr, Value]()
  def get($\rho$: Rep[Env], x: Ident): Rep[Addr] = $\rho$(x)
  def put($\rho$: Rep[Env], x: Ident, a: Rep[Addr]): 
    Rep[Env] = $\rho$ + (lift(x) -> a)
  def get($\sigma$: Rep[Store], a: Rep[Addr]): Rep[Value] = 
    $\sigma$.getOrElse(a, RepLattice[Value].bot)
  def put($\sigma$: Rep[Store], a: Rep[Addr], v: Rep[Value]): 
    Rep[Store] = $\sigma$ + (a -> RepLattice[Value].$\sqcup$(v, get($\sigma$, a)))
  def alloc($\sigma$: Rep[Store], x: Ident): Rep[Addr] = Addr(x)
  def num(i: Lit): Rep[Value] = Set(NumV())
  def branch0(cnd: Rep[Value], thn: => Ans, els: => Ans): Ans =
    thn $\sqcup$ els
  def prim_eval(op: Symbol, 
                v1: Rep[Value], v2: Rep[Value]): Rep[Value] = 
    Set(NumV())
  ...
}
\end{lstlisting}

@@

\begin{lstlisting}
def close(ev: EvalFun)($\lambda$: Lam, $\rho$: Rep[Env]): Rep[Value] = {
  val Lam(x, e) = $\lambda$
  val f: Rep[(Value, Store)]=>Rep[(Value,Store)] = {
    case (args: Rep[Value], $\sigma$: Rep[Store]) =>
      val args = as._1; val $\sigma$ = as._2; val $\alpha$ = alloc($\sigma$, x)
      ev(e, put($\rho$, x, $\alpha$), put($\sigma$, $\alpha$, args))
    }
  Set[AbsValue](CompiledClo(fun(f)))
}
def apply_closure(ev: EvalFun)
  (f: Rep[Value], arg: Rep[Value], $\sigma$: Rep[Store]): Ans = {
    reflectEffect(ApplyClosure(f, arg, $\sigma$))
  }
\end{lstlisting}

When generating code for an application, we can not directly apply the callee.
Instead, we emit code that calls a next-stage function @apply_closures_norep@.
As its unstaged counterpart, function @apply_closures_norep@
non-deterministically applies multiple target closures with the argument and
latest store, and finally returns the joined value and a single store. We
provide the definition of @apply_closures_norep@ in the runtime supporting code.

\begin{lstlisting}
case ApplyClosures(fs, arg, $\sigma$) =>
  emitValDef(sym, "apply_closures_norep(" + 
                  quote(fs) + "," + quote(arg) + 
                  "," + quote($\sigma$) + ")")
\end{lstlisting}

\subsection{Staged Fixpoint Iteration}

@@

\begin{lstlisting}
def iter(e: Expr, $\rho$: Rep[Env], $\sigma$: Rep[Store]): 
Rep[(Value,Store)] = {
  def iter_aux: Rep[Unit => (Value,Store)] = fun { () =>
    in = out; out = Map[Config, (Value,Store)]()
    cached_ev(e, $\rho$, $\sigma$)
    if (in === out) out((unit(e), $\rho$, $\sigma$)) else iter_aux()
  }
  iter_aux() // generated code that invokes iter_aux()
}
\end{lstlisting}

@@

\begin{lstlisting}
def cached_ev(e: Expr, $\rho$: Rep[Env], $\sigma$: Rep[Store]): 
Rep[(Value, Store)] = {
  val cfg: Rep[Config] = (unit(e), $\rho$, $\sigma$)
  if (out.contains(cfg)) { out(cfg) }
  else {
    val ans0 = in.getOrElse(cfg, RepLattice[(Value, Store)].bot)
    out = out + (cfg -> ans0)
    val ans1 = evev(cached_ev)(e, $\rho$, $\sigma$)
    out = out + (cfg -> (ans0 $\sqcup$ ans1)); ans1
  }
}
\end{lstlisting}

@@

Our staging schema works well in theory, but would suffer from code explosion,
and possibly runtime GC overhead when analyzing (specializing) large programs.
We present several optimizations that largely mitigate these issues. Still,
implementing these optimizations do not need to change the generic interpreter.

\paragraph{Specialized Data Structures}

@@

\paragraph{Selective Caching} It is observed that the two-fold caching is
wrapped for recursive call of our abstract interpreter. But this is even more
expensive when generating code for atomic expressions such as literal numbers or
variables. \todo{caching prevents nontermination}

@@

For example, to fold a singleton list, e.g., @List(x).foldLeft(init)(f)@ where
@x@ and @init@ are staged values, a naive code generator would faithfully
apply @foldLeft@ to the list with an anonymous function. But we can also utilize
the algebraic property of @foldLeft@ to generate cheaper code:
@List(x).foldLeft(init)(f) = f(init, x)@.
Since the function @f@ is known at current stage, we completely eliminate the
fold operation and function application. We apply several rewritings enabled by
partial-static data structures, such as @List@ and @Map@, which greatly reduce
the size of residual programs.

\paragraph{Heterogeneous Staging} The generated program is in A-Normal form,

@@

What we implemented is a big-step, compositional abstract interpreter in monadic
style, where \textit{compositional} means that every recursive call of our abstract
interpreter is applied to proper substructures of the current syntactic
parameters \cite{10.1007/3-540-61580-6_11}. This compositionality ensures that
specialization can be done by unfolding, as well as the termination of
specialization procedure. It is also possible to specialize small-step
operational abstract semantics through abstract compilation
\cite{Boucher:1996:ACN:647473.727587} --
\citet{Johnson:2013:OAA:2500365.2500604} implemented it for
optimizating Abstract Abstract Machines. However, the generated abstract
bytecode still requires another small-step abstract machine to execute, which is
an additional engineering efforts.

>>>>

In the previous sections, we have seen an unstaged abstract interpreter and a
staged concrete interpreter, now we begin describing the implementation of their
confluence -- a staged abstract interpreter.
Unsurprisingly, the staged abstract interpreter in this section has the same
abstract semantics with the unstaged version from Section~\ref{unstaged_abs}.
The approach from unstaged to staged abstract interpreter is modular,
principled, and does not sacrifice soundness or precision. The designer of the
analyzer has no need to rewrite the analysis. We first present the staged
lattices and staged monads, and then the staged version of primitive operations,
especially @close@, @ap_clo@ and @fix@, at last we discuss several optimizations
and our design.

@@

In Section \ref{stagedpoly_lat}, we exploited the higher-kinded type @R@ to
achieve stage polymorphism. Now we instantiate the type @R@ to @Rep@ and
still use power sets as an example to present its staged version.

\begin{lstlisting}
  trait RepLattice[S] extends Lattice[S, Rep]
  def RepSetLattice[T]: RepLattice[Set[T]] = new RepLattice[Set[T]] {
    lazy val ⊥: Rep[Set[T]] = Set[T]()
    lazy val ⊤: Set[T] = error("No representation for ⊤")
    def ⊑(l1: Rep[Set[T]], l2: Rep[Set[T]]): Rep[Boolean] = l1 subsetOf  l2
    def ⊔(l1: Rep[Set[T]], l2: Rep[Set[T]]): Rep[Set[T]]  = l1 union     l2
    def ⊓(l1: Rep[Set[T]], l2: Rep[Set[T]]): Rep[Set[T]]  = l1 intersect l2
  }
\end{lstlisting}

As we can see, the @RepSetLattice@ is an instance of @RepLattice@, which is also
an abstraction of the manipulations on the foundamental data structure. For
staged versions, these operations eventually are delegated to the @subsetOf@,
@union@ and @intersect@ operation on @Rep[Set[T]]@. We even don't have to change
the implementation code, but just the types – from @Set[T]@ to @Rep[Set[T]]@. In
the code generation part, these operations emit their corresponding next-stage
code. Again, other lattice structures such as products and maps are implemented
in a similar way.

@@

Now we implement the staged abstract semantics, where @R@ is instantiated as
@Rep@, and the abstract components are shared from the unstaged version.

\paragraph{Staged Monads for Abstract Interpretation}
We use the same moand stack structure as in the unstaged abstract interpreter,
but replacing \textit{all} the transformers to their staged version. The
following code shows this change. For the readability, we squeeze the three
inner transformers into a single monad @RepListReaderStateM[A, B, C]@, where @A@
and @B@ are both @Cache@ when used in @AnsM@. Now the result type is a pair of
two staged value: @Rep[List[(Value, Store)]]@ and @Rep[Cache]@.

\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type RepListReaderStateM[A, B, C] = RepListT[RepReaderT[RepStateT[RepIdM, A, ?], B, ?], C]
    type AnsM[T] = ReaderT[StateT[RepListReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[List[(Value, Store)]], Rep[Cache])
    ...
  }
\end{lstlisting}

Since our monad stack is uniformly using staged data, so the staged @ListT@ now
stores a staged value of type @Rep[List[A]]@, inside of a staged monad @M@.
\todo{staged ListT implementation}

\begin{lstlisting}
  case class ListT[M[_]: RepMonad, A](run: M[List[A]]) {
    def flatMap[B: Manifest](f: Rep[A] => ListT[M, B]): ListT[M, B] = ...
  }
\end{lstlisting}

\paragraph{Primitive Operations}

@@

\begin{lstlisting}
  def close(ev: EvalFun)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(List[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x); val ρ_* = ρ + (unit(x) → α); val σ_* = σ ⊔ Map(α → arg)
        ev(e)(ρ_*)(σ_*)(in)(out)
    }
    Set[AbsValue](emit_compiled_clo(f, λ, ρ))
  }
\end{lstlisting}

\begin{lstlisting}
  def lift_nd[T](vs: Rep[List[T]]): AnsM[T]
  def ap_clo(ev: EvalFun)(rator: Rep[Value], rand: Rep[Value]): Ans = for {
    clo <- lift_nd[AbsValue](rator.toList)
    σ <- get_store; in <- ask_in_cache; out <- get_out_cache
    val res: Rep[(List[(Value, Store)], Cache)] = emit_ap_clo(clo, rand, σ, in, out)
    _ <- put_out_cache(res._2)
    vs <- lift_nd[(Value, Store)](res._1)
    _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

\paragraph{Staged Caching and Fixpoint Iteration}

@@

\begin{figure}[t!]
  \centering
\begin{lstlisting}
  def fix(ev: EvalFun => EvalFun): EvalFun = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache
    val cfg: Rep[Config] = (unit(e), ρ, σ)
    val res: Rep[(List[(Value, Store)], Cache)] =
      if (out.contains(cfg)) (out(cfg).toList, out) // a next-stage if expression
      else { val m: Ans = for {
               _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
               v <- ev(fix(ev))(e)
               σ <- get_store
               _ <- update_out_cache(cfg, (v, σ))
             } yield v
             m(ρ)(σ)(in)(out) }
    _ <- put_out_cache(res._2); vs <- lift_nd(res._1); _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}
\vspace{-1em}
\caption{The staged co-inductive caching algorithm.}
\label{fig:staged_coind_cache}
\end{figure}

@@

\begin{lstlisting}
\end{lstlisting}

\paragraph{Code Generation} Code generation

@@

Our staging schema works well in theory, but in practice would suffer from code
explosion, and possibly runtime GC overhead when analyzing (specializing) large
programs. In this section, we present several optimizations that largely
mitigate these issues. Still, implementing these optimizations do not need to
change the generic interpreter.

\paragraph{Specialized Data Structures} \todo{need revise}

@@

\paragraph{Selective Caching} It is observed that the two-fold co-inductive
caching is used for every recursive call of our abstract interpreter. But this
is not necessary and even very expensive when generating code for atomic
expressions such as literals or variables, as they always terminate. Borrowing
the idea from the partition in ANF \cite{Flanagan:1993:ECC:155090.155113}, we
may use a selective caching algorithm:

\begin{lstlisting}
  def fix_select: EvalFun = e => e match {
    case Lit(_) | Var(_) | Lam(_, _) => eval(fix_select)(e)
    case _ => fix_cache(e)
  }
\end{lstlisting}

@@

Our treatment to binding-times is coarse-grained: @Expr@s is static, the rest of
the world are all dynamic. But this is not always true, since the static data
has to be used somewhere with the dyanmic operation. Partial-static data is a
way to improve binding-times and optimize the generated code.

For example, to fold a singleton list (often appears in @ListT@), e.g.,
@List(x).foldLeft(init)(f)@ where @x@ and @init@ are staged values, a naive code
generator would faithfully apply @foldLeft@ to the list with an anonymous
function. But we can also utilize the algebraic property of @foldLeft@ to
generate cheaper code: @List(x).foldLeft(init)(f) = f(init, x)@. Since the
function @f@ is known at current stage, we completely eliminate the fold
operation and function application. We apply several rewritings enabled by
partial-static data structures, such as @List@ and @Map@, which greatly reduces
the size of residual programs.

\paragraph{Heterogeneous Staging} The generated program is in A-Normal form. \todo{TODO}

@@

What we implemented is a big-step, compositional abstract interpreter in monadic
style, where \textit{compositional} means that every recursive call of our abstract
interpreter is applied to proper substructures of the current syntactic
parameters \cite{10.1007/3-540-61580-6_11}. This compositionality ensures that
specialization can be done by unfolding, as well as guarantees the termination
of specialization procedure. It is also possible to specialize small-step
operational abstract semantics through abstract compilation
\cite{Boucher:1996:ACN:647473.727587} -- as
\citet{Johnson:2013:OAA:2500365.2500604} implemented it for
optimizating Abstract Abstract Machines. However, the generated abstract
bytecode still requires another small-step abstract machine to execute, which is
an additional engineering efforts.
